Field-informed Reinforcement Learning for Learning Large-Scale Collective Tasks

### 2023-05-12 paper.tex
chore: add root in scala def

\and
%\linebreakand %\and
\IEEEauthorblockN{Mirko Viroli}
\IEEEauthorblockA{
%\textit{Department of Computer Science and Engineering} \\
\textit{%Alma Mater Studiorum--
University of Bologna}\\
Cesena, Italy\\
mirko.viroli@unibo.it % 0000−0003−2702−5702
}
\and
\IEEEauthorblockN{Lukas Esterle}
\IEEEauthorblockA{
%\textit{Department of Computer Science and Engineering} \\
\textit{%Alma Mater Studiorum--
Aarhus University}\\
Aarhus, Denmark\\
lukas.esterle@ece.au.dk}
}
%
\maketitle              % typeset the header of the contribution
%

%%% Comment command, to be removed before submission
\newcommand{\meta}[3]{\textcolor{#1}{\textbf{#2}: #3}}
\newcommand{\ga}[1]{\meta{red}{GA}{#1}}
\newcommand{\lukas}[1]{\meta{purple}{Lukas}{#1}}
\newcommand{\mv}[1]{\meta{green}{MV}{#1}}
%% ACRONYMS
\acrodef{DecPOMDP}[DecPOMDP]{decentralized partially observable Markov decision processes}
\acrodef{RL}[RL]{reinforcement learning}
\acrodef{MARL}[MARL]{multi-agent reinforcement learning}
\acrodef{MAARL}[ManyRL]{many-agent reinforcement learning}
\acrodef{MDP}[MDP]{Markov Decision Process}
\acrodef{MLP}[MLP]{multi-layer perceptron}
\acrodef{GNN}[GNN]{graph neural network}
\acrodef{CNN}[CNN]{convolutional neural network}
\acrodef{CTDE}[CTDE]{centralized training and decentralized execution}
\acrodef{DQN}[DQN]{Deep Q-Learning}
\ga{Page limit: 10 pages (include refs)}

@@

\section{Background and Motivation}
\label{sec:background}
In this section, we aim to provide a concise overview of the system we intend to address, 
 starting with a high-level system model. 
%
We then proceed to discuss relevant research areas that inform our approach. 
%
Finally, we present a formalization of the problem and a discussion of the motivations behind our proposed work.
\ga{Page budget: 1/2 page \\}
\ga{Plan: discussion about the problem of coordination in multi-agent systems 
and the need for a scalable approach. In doing this, we will discuss some of the existing approaches (declarative, RL and )}
\subsection{Swarm systems}
This article studies intelligent \emph{collective behaviours} within the context of \emph{large-scale} distributed systems.
 Specifically, we focus on \emph{Cyber-Physical Swarms} or \emph{swarm-like systems}, i.e.,  systems consisting of \emph{devices} (or nodes) equipped with \emph{sensors} and \emph{actuator} able to interact with the real world.
 Each of these nodes can interact with its direct neighbours, either based on physical (e.g. communication range) or logical distance (e.g. 1-hop neighbourhood).

%
Each node is governed by an autonomous software agent, executing a \emph{local} control loop. In each iteration of this loop, the agent can access the information available within its own \emph{context}. This context is comprised by information acquired by the node directly through its sensors and the information received from its neighbours. 
%
At the end of each control loop iteration, the node can send messages to its neighbourhood. This message may contain raw sensor data or aggregated information over time or from other neighbours.
% about its current evaluation and update its internal state.
%
%This general model can be adapted to several different scenarios, such as a network of sensors, a network of robots (e.g., swarm robotics), or a network of mobile phones.
Examples of such large-scale system can be a network of robots (e.g., swarm robotics), camera and IoT networks, or a network of mobile phones.
%
Our goal is to find a \emph{homogeneously} distributed controller $\pi$ (i.e., the same controller for the whole system) that, 
 starting from \emph{only} local configurations, leads 
 the system to achieve a certain \emph{collective} requirements through \emph{cooperation}, 
 such as spatial area coverage, phenomena tracking, and node aggregation.
%
The homogeneity of the controller is a key requirement to ensure the \emph{scalability} of the approach, 
 as it allows us to avoid the need for a centralised controller that would be a bottleneck for the system,
 and it is the typical choice in swarm-like systems~\cite{brambilla2013swarm,yang2021many,pmlr-v80-yang18d,DBLP:conf/aaai/ZhengYCZZWY18}.
%
\subsection{Field-based Coordination}
\ga{discussion about field used as a coordination mechanism in multi-agent systems. Introduction to AC}
The field-based coordination approaches utilise a concept of \emph{computational fields} 
 (or simply \emph{fields}), 
 which are \emph{distributed} data structures that associate locations with values and evolve over time. 
%
\emph{Aggregate computing}~\cite{Beal2015Computer}, a modern field-coordination approach, is rooted in earlier work on \emph{artificial potential fields}~\cite{DBLP:conf/icra/Warren89} and \emph{co-fields}~\cite{DBLP:journals/pervasive/MameiZL04}. 
% 
This \emph{macro-programming}~\cite{DBLP:journals/corr/abs-2201-03473} paradigm can be realised to devise collective and self-organizing behaviour through a composition of functions operating on fields. 
%
The fields map a set of individual agents to computational values, 
 allowing them to associate what they sense, process, and actuate. 
%
Fields are computed locally but subject to a global viewpoint, 
 enabling emergent collective behaviour through the interplay of the \emph{system model} (i.e., structure and dynamics of the systems)
 and the \emph{programming model} (i.e., how collective behaviours are expressed) known as field calculus~\cite{DBLP:conf/esocc/ViroliDB13,DBLP:journals/corr/ViroliADPB16,DBLP:journals/tocl/AudritoVDPB19}.
%
The foster is structurally similar to \emph{swarm-like systems}, 
 as it is composed of a set of \emph{agents} 
 that interact with each other through a \emph{network} of \emph{neighbourhood relations}.
%
The dynamics of each agent follow a local control loop called \emph{round}
 that comprises the following steps:
\begin{enumerate}
  \item \emph{Sense}: the agent acquires information from the environment and collects the messages of its neighbourhood, building a \emph{context};
  \item \emph{Compute}: the agent executes a \emph{program} that computes an \emph{export}, that is a set of values to be sent to the neighbourhood;
  \item \emph{Act}: the agent sends the export to its neighbourhood and updates its internal state acting on the environment.
\end{enumerate}
The proactive and iterative execution of the round by each agent leads to the emergence of collective behaviour specified by the program.

The programming model is ruled by the \emph{field calculus}~\cite{DBLP:conf/esocc/ViroliDB13,DBLP:journals/corr/ViroliADPB16,DBLP:journals/tocl/AudritoVDPB19}, 
 a core language that allows the expression of collective behaviours through the composition of \emph{functions} operating on fields.

\lukas{still not sure how the coordination works or if it has been explained in the above paragraph.}
\subsection{Graph Neural Networks}
%Over the years, several different neural networks have been proposed to solve specific tasks, ranging from 
%  simple \ac{MLP}~\cite{werbos1975beyond} to \ac{CNN}~\cite{oshea2015introduction}.
%  \lukas{more examples? }
%%
%Whereas \ac{CNN} is designed to be used with image and spatial-like data (e.g., audio), 
 \ac{GNN} is a novel neural network model used to process graph-structured data with deep learning approaches.
%
Let $G = (V, E)$ a graph  
where $E$ defines the neighbourhood relations for each participating node, 
and $V$ identifies the nodes present in the graph. %at that time.  \lukas{no time in the definition}
% 
Each node $v \in V$ is associated with an observation (or feature set) $o_v$.
\lukas{Over time this graph $G$ can change and therefore we define $G_t$ as the graph $G$ at a specific time $t$.} 
%
Given $o_v$, %a feature vector associated with each node $v \in V$, 
 the goal of a \ac{GNN} is to learn the node embedding $h_v$ for each node $v \in V$. The node embedding $h_v$ describes the node in the network and summarises the geometric properties of the graph in this location, allowing for comparison of various nodes in the graph.
 In modern \acp{GNN}, the node embedding $h_v$ is computed by aggregating information from the node's neighbours $N(v)$,
  and then combining it with the node's current embedding $h_v$.
 The graph is further partitioned into layers with the node $v$ at its bottom layer $0$. With each traversal of an edge $e \in E$ from $v$, the number of layers increased
Formally, a \ac{GNN} can be defined by two phases:
\begin{equation}
a_{v}^{(k)}=\operatorname{AGGREGATE}^{(k)}\left(\left\{h_{u}^{(k-1)}: u \in \mathcal{N}(v)\right\}\right)   
\end{equation}
\begin{equation}
h_{v}^{(k)}=\operatorname{COMBINE}^{(k)}\left(h_{v}^{(k-1)}, a_{v}^{(k)}\right)
\end{equation}
where $h_{v}^{k}$ is the embedding of node $v$ within the $k$-th layer ($h_v^0 = o_v$), 
 $\mathcal{N}(v)$ is the set of neighbours of node $v$ computed from $E$.
%
The differential part comes into play in the $\operatorname{COMBINE}$ function, 
 which is usually a differentiable function such as a neural network layer.
%
$\operatorname{AGGREGATE} $ instead is a function that aggregates the information from the neighbours of a node $v$
  and it could be a simple sum, max or sum of products. More complex aggregation functions are available~\cite{pellegrini2020learning}.
%
This formulation allows GNNs to effectively process and extract features from graph-structured data by iteratively aggregating and transforming information from the node's neighbours.
\lukas{do we need definitions of $o_v$, $h_v^*$ over time $t$?}

\acp{GNN} are used in several application areas such as social network analysis, chemistry, and physics.
In this paper, we use GNNs to learn a local behaviour for each agent in a multi-agent system (more details in Section~\ref{sec:approach}).
\subsection{Many-Agent Reinforcement Learning}
\ga{plan: we could discuss briefly about standard approach (few agents), and mean-field approach (many agents). 
This will lead to our approach, which is a combination of DQN (or any Deep RL approach) and GNN.}

@@

\ga{Plan: we should discuss the approach in detail, including the system model, and the learning dynamics.
Particularly, I would like to highlight how each node in the graph is a local controller,
but it could be seen as a global evolution, therefore we can use global information to inform the local controller.}

@@

\ga{Plan: 
  discussion about the typical AC evaluation model, specifying how GNN (with 1-hop information diffusion)
  can be easily integrated with this distributed model

@@

\ga{
  Plan: discuss the chosen approach, that is Centralised Training and Distributed Execution.
  Particularly, we use DQN in which the neural network is a GNN. So, the dataset consists in a set of graphs,
  where each graph is a snapshot of the system. 
  The target is the Q-value of the action taken in the current state.
  Explain how this can be then used for distributed controllers.

@@

\section{Evaluation}
\label{sec:eval}
\ga{Page budget: 2 pages \\}
\ga{Plan: we should discuss the robot aggregation scenario, specifying the state space, the action space, and the reward function.
I dunno if it makes sense to discuss the variants (i.e., fixed area, two areas and moving area). We will see.
}
To test the effectiveness of the proposed approach, 
 we decided to experiment with a case study related to swarm robotics, 
 specifically the tracking and coverage of a spatio-temporal phenomenon, cf. tracking a wildfire 
 or monitoring the water levels in a canal.

>>>>

\and
%\linebreakand %\and
\IEEEauthorblockN{Mirko Viroli}
\IEEEauthorblockA{
%\textit{Department of Computer Science and Engineering} \\
\textit{%Alma Mater Studiorum--
University of Bologna}\\
Cesena, Italy\\
mirko.viroli@unibo.it % 0000−0003−2702−5702
}
\and
\IEEEauthorblockN{Lukas Esterle}
\IEEEauthorblockA{
%\textit{Department of Computer Science and Engineering} \\
\textit{%Alma Mater Studiorum--
Aarhus University}\\
Aarhus, Denmark\\
lukas.esterle@ece.au.dk}
}
%
\maketitle              % typeset the header of the contribution
%
\input{scala.tex}
\lstset{language={scafi}}
%%% Comment command, to be removed before submission
\newcommand{\meta}[3]{\textcolor{#1}{\textbf{#2}: #3}}
\newcommand{\ga}[1]{\meta{red}{GA}{#1}}
\newcommand{\lukas}[1]{\meta{purple}{Lukas}{#1}}
\newcommand{\mv}[1]{\meta{green}{MV}{#1}}
%% ACRONYMS
\acrodef{DecPOMDP}[DecPOMDP]{decentralized partially observable Markov decision processes}
\acrodef{RL}[RL]{reinforcement learning}
\acrodef{MARL}[MARL]{multi-agent reinforcement learning}
\acrodef{MAARL}[ManyRL]{many-agent reinforcement learning}
\acrodef{MDP}[MDP]{Markov Decision Process}
\acrodef{MLP}[MLP]{multi-layer perceptron}
\acrodef{GNN}[GNN]{graph neural network}
\acrodef{CNN}[CNN]{convolutional neural network}
\acrodef{CTDE}[CTDE]{centralized training and decentralized execution}
\acrodef{DQN}[DQN]{Deep Q-Learning}
\ga{Page limit: 10 pages (include refs)}

@@

\section{Background and Motivation}
\label{sec:background}
In this section, we aim to provide a concise overview of the system we intend to address, 
 starting with a high-level system model. 
%
We then proceed to discuss relevant research areas that inform our approach. 
%
Finally, we present a formalization of the problem and a discussion of the motivations behind our proposed work.
\ga{Page budget: 1/2 page \\}
%\ga{Plan: discussion about the problem of coordination in multi-agent systems and the need for a scalable approach. In doing this, we will discuss some of the existing approaches (declarative, RL and )}
\subsection{Swarm systems}
This article studies intelligent \emph{collective behaviours} within the context of \emph{large-scale} distributed systems.
 Specifically, we focus on \emph{Cyber-Physical Swarms} or \emph{swarm-like systems}, i.e.,  systems consisting of \emph{devices} (or nodes) equipped with \emph{sensors} and \emph{actuator} able to interact with the real world.
 Each of these nodes can interact with its direct neighbours, either based on physical (e.g. communication range) or logical distance (e.g. 1-hop neighbourhood).

%
Each node is governed by an autonomous software agent, executing a \emph{local} control loop. In each iteration of this loop, the agent can access the information available within its own \emph{context}. This context is comprised by information acquired by the node directly through its sensors and the information received from its neighbours. 
%
At the end of each control loop iteration, the node can send messages to its neighbourhood. This message may contain raw sensor data or aggregated information over time or from other neighbours.
% about its current evaluation and update its internal state.
%
%This general model can be adapted to several different scenarios, such as a network of sensors, a network of robots (e.g., swarm robotics), or a network of mobile phones.
Examples of such large-scale system can be a network of robots (e.g., swarm robotics), camera and IoT networks, or a network of mobile phones.
%
Our goal is to find a \emph{homogeneously} distributed controller $\pi$ (i.e., the same controller for the whole system) that, 
 starting from \emph{only} local configurations, leads 
 the system to achieve a certain \emph{collective} requirements through \emph{cooperation}, 
 such as spatial area coverage, phenomena tracking, and node aggregation.
%
The homogeneity of the controller is a key requirement to ensure the \emph{scalability} of the approach, 
 as it allows us to avoid the need for a centralised controller that would be a bottleneck for the system,
 and it is the typical choice in swarm-like systems~\cite{brambilla2013swarm,yang2021many,pmlr-v80-yang18d,DBLP:conf/aaai/ZhengYCZZWY18}.
%
\subsection{Field-based Coordination}
%\ga{discussion about field used as a coordination mechanism in multi-agent systems. Introduction to AC}
The field-based coordination approaches utilise a concept of \emph{computational fields} 
 (or simply \emph{fields}), 
 which are \emph{distributed} data structures that associate locations with values and evolve over time. 
%
\emph{Aggregate computing}~\cite{Beal2015Computer}, a modern field-coordination approach, is rooted in earlier work on \emph{artificial potential fields}~\cite{DBLP:conf/icra/Warren89} and \emph{co-fields}~\cite{DBLP:journals/pervasive/MameiZL04}. 
% 
This \emph{macro-programming}~\cite{DBLP:journals/corr/abs-2201-03473} paradigm can be realised to devise collective and self-organizing behaviour through a composition of functions operating on fields. 
%
The fields map a set of individual agents to computational values, 
 allowing them to associate what they sense, process, and actuate. 
%
Fields are computed locally but subject to a global viewpoint, 
 enabling emergent collective behaviour through the interplay of the \emph{system model} (i.e., structure and dynamics of the systems)
 and the \emph{programming model} (i.e., how collective behaviours are expressed) known as field calculus~\cite{DBLP:conf/esocc/ViroliDB13,DBLP:journals/corr/ViroliADPB16,DBLP:journals/tocl/AudritoVDPB19}.
%
The foster is structurally similar to \emph{swarm-like systems}, 
 as it is composed of a set of \emph{agents} 
 that interact with each other through a \emph{network} of \emph{neighbourhood relations}.
%
The dynamics of each agent follow a local control loop called \emph{round}
 that comprises the following steps:
\begin{enumerate}
  \item \emph{Sense}: the agent acquires information from the environment and collects the messages of its neighbourhood, building a \emph{context};
  \item \emph{Compute}: the agent executes a \emph{program} that computes an \emph{export}, that is a set of values to be sent to the neighbourhood;
  \item \emph{Act}: the agent sends the export to its neighbourhood and updates its internal state acting on the environment.
\end{enumerate}
The proactive and iterative execution of the round by each agent leads to the emergence of collective behaviour specified by the program.

The programming model is ruled by the \emph{field calculus}~\cite{DBLP:conf/esocc/ViroliDB13,DBLP:journals/corr/ViroliADPB16,DBLP:journals/tocl/AudritoVDPB19}, 
 a core language that allows the expression of collective behaviours through the composition of \emph{functions} operating on fields.
%
Field calculus includes main operators for expressing \emph{spatio-temporal} compotations that allow for 
\emph{i)} the progression of values over time, achieved by transforming a value computed in a previous round into a new value; 
\emph{ii)} the exchange of data with neighbouring fields, where received data is validated by adjacent fields; and 
\emph{iii)} the conditional division of computation into distinct domains of collective computation. 
%
We avoid presenting here the full details of these operators since they are not essential to the contribution. 
For further elaboration on the actual field calculus, please refer to the work of~\cite{DBLP:journals/jlap/ViroliBDACP19}.

On these minimal operators, 
 it is then possible to build \emph{self-organizing} coordination blocks.
 One of the founding blocks is the \emph{gradient} essential for 
 \emph{information} flows~\cite{DBLP:conf/saso/WolfH07}.
%
This operator generates a numerical field that represents the minimum distance from a source zone. 
%
In other words, it maps a Boolean field (where ``true'' indicates the presence of a source node and ``false'' indicates its absence) to a distance field that indicates the proximity to the nearest source. In ScaFi (a scala implementation of aggregate computing) it is simply expressed as:
\begin{lstlisting}
def gradient(source: Boolean): Double
\end{lstlisting}
Along the gradient, it is possible to cast information within the system, accumulating value during the programming. This pattern is called \emph{gradient-cast} (or \emph{G}) 
\begin{lstlisting}
def G[V](source: Boolean, value: V, acc: V => V): V
\end{lstlisting}
\ga{more description here}
this is used in the use case to accumulate the distance from the phenomena that is tracked
\lukas{still not sure how the coordination works or if it has been explained in the above paragraph.}
\subsection{Graph Neural Networks}
%Over the years, several different neural networks have been proposed to solve specific tasks, ranging from 
%  simple \ac{MLP}~\cite{werbos1975beyond} to \ac{CNN}~\cite{oshea2015introduction}.
%  \lukas{more examples? }
%%
%Whereas \ac{CNN} is designed to be used with image and spatial-like data (e.g., audio), 
 \ac{GNN} is a novel neural network model used to process graph-structured data with deep learning approaches.
%
Let $G = (V, E)$ a graph  
where $E$ defines the neighbourhood relations for each participating node, 
and $V$ identifies the nodes present in the graph. %at that time.  \lukas{no time in the definition}
% 
Each node $v \in V$ is associated with an observation (or feature set) $o_v$.
\lukas{Over time this graph $G$ can change and therefore we define $G_t$ as the graph $G$ at a specific time $t$.} 
%
Given $o_v$, %a feature vector associated with each node $v \in V$, 
 the goal of a \ac{GNN} is to learn the node embedding $h_v$ for each node $v \in V$. The node embedding $h_v$ describes the node in the network and summarises the geometric properties of the graph in this location, allowing for comparison of various nodes in the graph.
 In modern \acp{GNN}, the node embedding $h_v$ is computed by aggregating information from the node's neighbours $N(v)$,
  and then combining it with the node's current embedding $h_v$.
 The graph is further partitioned into layers with the node $v$ at its bottom layer $0$. With each traversal of an edge $e \in E$ from $v$, the number of layers increased
Formally, a \ac{GNN} can be defined by two phases:
\begin{equation}
a_{v}^{(k)}=\operatorname{AGGREGATE}^{(k)}\left(\left\{h_{u}^{(k-1)}: u \in \mathcal{N}(v)\right\}\right)   
\end{equation}

\begin{equation}
h_{v}^{(k)}=\operatorname{COMBINE}^{(k)}\left(h_{v}^{(k-1)}, a_{v}^{(k)}\right)
\end{equation}
where $h_{v}^{k}$ is the embedding of node $v$ within the $k$-th layer ($h_v^0 = o_v$), 
 $\mathcal{N}(v)$ is the set of neighbours of node $v$ computed from $E$.
%
The differential part comes into play in the $\operatorname{COMBINE}$ function, 
 which is usually a differentiable function such as a neural network layer.
%
$\operatorname{AGGREGATE} $ instead is a function that aggregates the information from the neighbours of a node $v$
  and it could be a simple sum, max or sum of products. More complex aggregation functions are available~\cite{pellegrini2020learning}.
%
This formulation allows GNNs to effectively process and extract features from graph-structured data by iteratively aggregating and transforming information from the node's neighbours.
\lukas{do we need definitions of $o_v$, $h_v^*$ over time $t$?}

\acp{GNN} are used in several application areas such as social network analysis, chemistry, and physics.
In this paper, we use GNNs to learn a local behaviour for each agent in a multi-agent system (more details in Section~\ref{sec:approach}).
\subsection{Many-Agent Reinforcement Learning}
%\ga{plan: we could discuss briefly about standard approach (few agents), and mean-field approach (many agents). 
%This will lead to our approach, which is a combination of DQN (or any Deep RL approach) and GNN.}

@@

%\ga{Plan: we should discuss the approach in detail, including the system model, and the learning dynamics. Particularly, I would like to highlight how each node in the graph is a local controller, but it could be seen as a global evolution, therefore we can use global information to inform the local controller.}

@@

%\ga{Plan: 
%  discussion about the typical AC evaluation model, specifying how GNN (with 1-hop information diffusion)
%  can be easily integrated with this distributed model
%}

@@

%\ga{
%  Plan: discuss the chosen approach, that is Centralised Training and Distributed Execution.
%  Particularly, we use DQN in which the neural network is a GNN. So, the dataset consists in a set of graphs,
%  where each graph is a snapshot of the system. 
%  The target is the Q-value of the action taken in the current state.
%  Explain how this can be then used for distributed controllers.
%}

@@

\section{Evaluation}
\label{sec:eval}
\ga{Page budget: 2 pages \\}
%\ga{Plan: we should discuss the robot aggregation scenario, specifying the state space, the action space, and the reward function.
%I dunno if it makes sense to discuss the variants (i.e., fixed area, two areas and moving area). We will see.}
To test the effectiveness of the proposed approach, 
 we decided to experiment with a case study related to swarm robotics, 
 specifically the tracking and coverage of a spatio-temporal phenomenon, cf. tracking a wildfire 
 or monitoring the water levels in a canal.
